What We Eval in the Shadows

### 2021-04-14 paper/main.tex
origins

@@

\subsection{The origins of \eval}

Where does the expression passed to \eval come from? There are various means of
creating that expression, these are associated with particular use cases.

\begin{compactitem}[---]
\item {\it Reflection:} The \c{match.call} function reflectively captures
  the expression that called the current function.
\item {\it Constructed:} Expressions can be constructed by invoking the
  \c{quote}, \c{enquote}, \c{expression} and \c{substitute} functions. Function arguments are passed as promises, suspended computations, and  \c{substitute} is sometimes used to retrieve the source expression associated to the promise.
\item {\it String:} Finally, expressions can be created from strings by invoking
  one of \c{parse}, \c{str2expression} and \c{str2lang}.
\end{compactitem}

Table~\ref{tab:provenance} summarizes the expression provenance in our corpus.
This data is obtained by dynamically tainting values as they are produced by the
various sources, due to technical reasons there is some imprecision in the
results. Manual inspection of numerous examples suggest that ordering is correct.

\begin{table}[h]\small
\begin{tabular}{r|r|r} \hline
Origin  & \#sites & \%sites \\\hline
Constructed & \packageNbConstructedSites & \packageNbConstructedSitePercent \\ \hline
String & \packageNbStringSites & \packageNbStringSitePercent \\ \hline
Reflection &  \packageNbMatchCallExprsSites & \packageMatchCallExprsSitePercent  \\\hline
Other & \packageNbUnknownSites & \packageNbUnknownSitePercent \\\hline
\end{tabular}
\caption{Provenance}\label{tab:provenance}
\end{table}

The string origin can sometimes be correlated with dynamic code loading. This is
what the base functions \c{source} and \c{sys.source} do. We observed few calls
(\packageNbParseFromFileSites in total) that consume the result of calling
\c{parse} on a file. Most of the calls build strings programmatically. We also
identified one function \c{invokeRestartInteractively} that prompts the user for
input, parses it, and passes it to \eval.  The use of strings seems to correlate
with less sophisticated programmers, as in our Kaggle corpus where
\kaggleParseExprsSitePercent of site use strings.

>>>>

\paragraph{Discussion}
The data presented in this section is what one could expect. ({\it Function})
The majority of \eval call sites access the environment of some function on the
call stack. Usually, the current function, but sometimes frames arbitrarily far
up the call stack. One thing the data does not say is how that environment was
obtained. The expected case is that the environment was the one obtained from a
promise; as promises combine code and the environment in which that code
originates. Likely less frequent are cases where \eval is provided the results
of programmatically selecting some call stack. ({\it Synthetic}) The relative
high frequency of synthetic environments corresponds to cases when one wants to
evaluate an expression in either a restricted environment or use a data
structure as an environment. ({\it Global}) The fact that global environment
show up so frequently is likely an artifact of the way the code is run. ({\it
  Package}) There are only 92 sites that use a package environment in \eval,
this is probably best as it is bad form to add bindings or modify functions
of a loaded package.

@@

\newpage
\subsection{The origins of \eval}

Where does the expression passed to \eval come from? There are various means of
creating that expression, these are associated with particular use cases, we
classify them in three categories:

\begin{compactitem}[---]
\item {\it Reflection:} This group corresponds to uses of the \c{match.call}
  function to reflectively capture the expression that invoked the current
  function.
\item {\it Constructed:} Expressions can be constructed by invoking the
  \c{quote}, \c{enquote}, \c{expression}. Function arguments are passed as
  promises and \c{substitute} is used to retrieve the source expression
  associated to the promise.
\item {\it String:} Finally, expressions can be created from strings by invoking
   \c{parse}, \c{str2expression} or \c{str2lang}.
\end{compactitem}

\noindent
Table~\ref{tab:provenance} summarizes the expression provenance in our corpus.
This data is obtained by dynamically tainting values as they are produced by the
various sources, due to technical reasons there is some imprecision in the
results. In particular, we are not able to classify all sites. Manual inspection
of numerous examples suggest that the classified results are accurate.

\begin{wraptable}{r}{5cm}\small\centering
\begin{tabular}{r|r|r} \hline
Origin  & \#sites & \%sites \\\hline
Constructed & \packageNbConstructedSites & \packageNbConstructedSitePercent \\
String & \packageNbStringSites & \packageNbStringSitePercent \\
Reflection &  \packageNbMatchCallExprsSites & \packageMatchCallExprsSitePercent\\\hline
\end{tabular}
\caption{Provenance}\label{tab:provenance}
\end{wraptable}

Strings could correlate with dynamic code loading. This is what the base
functions \c{source} and \c{sys.source} do. We observed few calls
(\packageNbParseFromFileSites in total) that consume the result of calling
\c{parse} on a file. Most of the calls build strings programmatically. We also
identified one function \c{invokeRestartInteractively} that prompts the user for
input, parses it, and passes it to \eval. The use of strings seems to correlate
with less sophisticated programmers; in our Kaggle corpus
\kaggleParseExprsSitePercent of site use strings.

\paragraph{Discussion}
The origin data suggests that constructing expressions from strings is a
minority of the use cases, instead the constructed category shows that the
majority of \evals comes code that was processed by the compiler, and may be
slightly modified by the programmer before invoking \eval. Both constructed and
reflection categories roughly correspond to meta-programming. Some of these use
cases could likely be replaced by macros if the R designers could be convinced
to overcome their distaste for those.
